```{r, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
```

[![CI](https://github.com/jonclayden/imply/actions/workflows/ci.yaml/badge.svg)](https://github.com/jonclayden/imply/actions/workflows/ci.yaml)

# imply: Efficient Generalised Images for R

`imply` (a portmanteau of "image" and "apply") provides data structures for
*generalised images* — two- or three-dimensional rasters that may hold a
vector or time series at each location rather than a single intensity —
together with infrastructure for memory-efficiently applying functions to
them. The primary use case is medical imaging, but nothing in the package
assumes it.

The package deliberately sits *below* the existing R imaging stack: it has no
dependency on any file format, and an image's geometry is a plain 4x4 affine.
Reading, writing and interoperating with other image classes is out of scope,
and left to packages that sit above this one.

`imply` is distinguished by

- three storage representations — an ordinary R array, a masked sparse form,
  and a narrow packed form — behind **one** apply engine, so a kernel written
  once runs unchanged whichever way the data happens to be held;
- `imapply()`, which gives bit-identical results to `base::apply()` but at
  roughly half the peak memory, because a sub-array is gathered directly
  through the stride vector rather than the whole array first being permuted
  into a fresh copy; and
- modest dependencies: only `Rcpp`, `S7`, `methods`, `parallel`, `stats` and
  `utils`.

The latest development version of the package can be installed from GitHub
using the `remotes` package.

```{r, eval=FALSE}
## install.packages("remotes")
remotes::install_github("jonclayden/imply")
```

## Image types

`imply` supports three storage representations, chosen for how the data is
best used rather than how it happens to arrive.

| | class | storage | good for |
|---|---|---|---|
| dense | `denseImage` | an ordinary R array, any storage mode | general use; anything not covered below |
| sparse | `sparseImage` | a bitset mask over spatial locations, plus values packed in location order | images that are mostly zero outside a mask, such as brain-masked data |
| packed | `packedImage` | a raw vector reinterpreted as `float32`/`int16`/…, with a slope and intercept | large images where the values are not all needed as `double` at once, such as raw scanner data |

A `denseImage` wraps an array with some geometry attached: how many of the
leading dimensions are spatial, the size of a voxel, and the 4x4 affine
mapping voxel to world coordinates.

```{r}
library(imply)

set.seed(1)
data <- array(rnorm(4 * 5 * 6 * 10), dim = c(4, 5, 6, 10))
image <- denseImage(data, voxelSize = c(2, 2, 2.5), spatial = 3)
image
```

Here the image is 4 x 5 x 6 spatially, with a ten-point time series at each
location — `spatial = 3` says so explicitly, though it would have been
inferred anyway, since `spatial()` defaults to the first three dimensions or
however many there are, whichever is fewer.

```{r}
dim(image)
spatial(image)
voxelSize(image)
worldTransform(image)
```

Voxel size and world placement are stored, and can be set, independently of
one another, so replacing one never has to touch the other — see `?geometry`.
`fromVoxel()` and `toVoxel()` convert points between voxel and world space
using the composed transform, in the zero-based voxel coordinates it uses (as
opposed to R's one-based array indexing):

```{r}
toVoxel(c(0, 0, 0), image)
fromVoxel(c(1, 1, 1), image)
```

Because a `denseImage` is a plain array underneath, ordinary R operations —
indexing, arithmetic, comparison — work on it directly and return a plain
array, since an arbitrary index or elementwise result has no well-defined
geometry of its own.

### Converting between representations

`asSparse()`, `asDense()` and `asPacked()` convert freely between the three
representations; `asDenseImage()` promotes any plain array to a `denseImage`
without touching data already in one.

```{r}
sparse <- asSparse(image > 2) # a location is "present" only while some value in it is
sparse
sparseness(sparse)
```

A sparse image's mask can be recovered as a logical array with `mask()`, and
its stored values as a matrix with one column per present location — the
shape most voxelwise analysis wants — with `maskedMatrix()`. Because a
`sparseImage` stores its values already packed that way, this costs no copy.

```{r}
mask(sparse)[1:2, 1:2, 1]
dim(maskedMatrix(sparse))
```

`asPacked()` narrows storage to one of the NIfTI-style integer or `float32`
types, choosing a slope and intercept automatically so the type's whole range
is used.

```{r}
packed <- asPacked(image, type = "int16")
packed
```

`asDense()` unpacks whichever compact representation it is given, so code
that just wants ordinary values need not ask which one it has:

```{r}
back <- asDense(packed)
range(as.array(back) - as.array(image)) # narrowing loses some precision, but very little
```

`storageType()` reports `"double"`, `"integer"` and so on for anything that
is not a packed image, and the packed type otherwise, so it can be used
without first checking which kind of image is in hand. `isDenseImage()`,
`isSparseImage()` and `isPackedImage()` (and the umbrella `isImage()`) are the
right way to test what an image is — because S7 qualifies its class names
with the package, `inherits(x, "denseImage")` is always `FALSE`.

```{r}
storageType(image)
storageType(packed)
isPackedImage(packed)
```

A `template = ` argument, accepted by all three constructors and by
`asDense()`/`asSparse()`/`asPacked()`, carries geometry over from an existing
image when only the data is changing.

```{r}
doubled <- denseImage(as.array(image) * 2, template = image)
voxelSize(doubled)
```

## Applying and reducing functions

`imapply()` is the memory-efficient analogue of `base::apply()`: it gives
the same answer for the same `margin`, but gathers each sub-array directly
through the image's strides rather than permuting the whole array into a
fresh copy first, so peak memory is the input plus the result rather than
twice the input plus the result.

```{r}
totals <- imapply(image, 4, sum)
totals
```

Three further verbs differ only in how much of the *space* is handed to the
function at a time — a single location, a line, or a slice — with the values
held at each location always travelling along with it:

```
dim(x) = 4 x 5 x 6 x 10, spatial = 3

voxelApply(x, f)           f sees  10
lineApply(x, f, axis = 1)  f sees  4 x 10
sliceApply(x, f, axis = 3) f sees  4 x 5 x 10
```

```{r}
means <- voxelApply(image, mean)
means # a single value per location comes back as an image, with the same geometry
```

```{r}
lineMeans <- lineApply(image, rowMeans, axis = 1)
dim(lineMeans)

sliceMeans <- sliceApply(image, mean, axis = 3)
sliceMeans
```

`voxelApply()` accepts a `mask = ` — a logical array, a sparse image, or any
numeric array where non-zero means selected — and visits only those
locations, leaving `fill` (`0` by default) everywhere else. The loop runs
over a packed matrix of just the selected columns, so a sparse image being
masked by its own mask costs nothing extra.

```{r}
masked <- voxelApply(image, mean, mask = sparse)
masked[1:2, 1:2, 1]
```

For a long-running call, `progress = TRUE` draws a text bar showing
percentage complete and throughput in voxels per second (or pass a function
of `(done, total)` to report however you like):

```{r, eval=FALSE}
voxelApply(image, slow_function, progress = TRUE)
#>   |======================            | 62%  1.4M voxels/s
```

### Reductions

A handful of summaries — `min`, `max`, `range`, `which.min`, `which.max`,
plus `sum`, `mean`, `prod`, `var`, `sd`, `any`, `all` and `countNA` — can be
computed without calling back into R at all, using `imreduce()`:

```{r}
imreduce(image, 4, "range")
```

`imapply()` recognises calls to the base equivalents of `min`, `max`,
`range`, `which.min` and `which.max` and routes to `imreduce()`
automatically, since these are the only ones provably guaranteed to give the
same answer either way — the arithmetic reductions can differ from
`imapply(x, margin, sum)` in the last bit, because R accumulates `sum()` and
`mean()` in long double.

```{r}
identical(imapply(image, 4, max), imreduce(image, 4, "max"))
```

## Parallelism

Work is parallelised in one of two ways, according to what is being run.
Compiled reductions run concurrently in-process, using Grand Central Dispatch
where it is available and OpenMP otherwise:

```{r}
parallelBackend()
```

An R function passed to `imapply()` and its variants cannot run on a worker
thread — R's interpreter is not reentrant — so that work is instead
parallelised by forking the R session, which is unavailable on Windows
(`canFork()` reports whether it is available here). Pass `threads = ` to any
apply or reduce call, or set it globally:

```{r, eval=FALSE}
options(imply.threads = 4)
```

```{r}
resolveThreads(2)
```

## API

The apply/reduce engine, the accessors that let one kernel body read dense,
sparse and narrowly-stored images alike, and the voxel-to-world geometry
machinery behind `worldTransform()` and friends, are all available to other
packages' compiled code — not just to `imply` itself. Everything under
`inst/include/` is header-only, so using it costs nothing to link against.

```
LinkingTo: imply
```

in a package's `DESCRIPTION` is enough; there is no library to link, only the
one header to include.

```c++
#include "imply.h"
```

The pieces most likely to be wanted from outside:

- **`Raster<D>`** — an n-dimensional index space with general strides, split
  at a runtime `spatial` index into leading spatial dimensions and trailing
  value dimensions. `FixedRaster<D>` keeps extents and indices on the stack
  when the dimensionality is known while compiling; `DynamicRaster` covers
  the rest.
- **`OffsetWalker`** — odometer traversal of an arbitrary subset of an
  image's dimensions, yielding memory offsets one addition at a time with no
  allocation. This is what lets a sub-array be gathered without permuting
  the whole image first.
- **`ImageSpace`** and **`Affine`** — voxel-to-world geometry, with the point
  conversions and rounding strategies that go with it, and no dependency on
  any file format.
- **`parallelFor`** — divides a range into chunks and runs them concurrently
  over libdispatch, OpenMP or a plain loop, whichever was compiled in.
  Nothing passed to it may touch the R API, allocate R objects, or draw from
  R's RNG, none of which are thread-safe.
- **`dispatchType`**, **`dispatchNarrowType`** and **`dispatchDims`** — the
  R-boundary switches from a runtime `SEXP`'s storage mode (or
  dimensionality) to a compile-time type tag, meant to be called exactly
  once per entry point so that everything below it is fully typed.
- **`DenseAccessor`**, **`SparseAccessor`** and **`NarrowAccessor`**, plus
  **`LocationMask`** — an `operator[]` over a plain pointer, a masked and
  packed array, or narrow storage respectively, sharing one interface so
  that a kernel written against it serves all three without modification.
- **`Sink`**, **`VectorSink`**, **`TypedSink`** and **`ListSink`** —
  preallocated result accumulation for a per-call loop, falling back to a
  list only once a result turns out not to fit a plain vector.

A kernel that sums a numeric vector, written once against the shared
accessor interface, therefore works whether the underlying storage is a
plain R vector or something more compact:

```c++
// [[Rcpp::depends(imply)]]
#include "imply.h"
#include <Rcpp.h>

using namespace imply;

// One kernel body serves dense, sparse and narrow storage alike; only the
// accessor differs
template <typename Accessor>
double sumOver (const Accessor &values, const Extent n)
{
    double total = 0;
    for (Extent i = 0; i < n; i++)
        total += static_cast<double>(values[i]);
    return total;
}

// [[Rcpp::export]]
double denseSum (Rcpp::NumericVector x)
{
    DenseAccessor<double> accessor(REAL(x));
    return sumOver(accessor, Rf_xlength(x));
}
```

There is no separate API reference beyond the headers themselves: each one
opens with a comment explaining what it is for, and `imply.h` is a short
umbrella summarising all of them together — reading it is the fastest way
in.
